---
title: Statistical analysis of the effect of environmental variables on abundance
  of flounder
author: "Olga Lyashevska"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  pdf_document:
    toc: true
fig_caption: true  # Enable figure captions
---

```{r setup}
# load packages 
packages <- c("ggplot2", "MASS", "rmarkdown", "tinytex", "reshape2", "glmmTMB", "DHARMa", "emmeans")
lapply(packages, library, character.only = TRUE)
knitr::opts_chunk$set(fig.path = "figure/", dev = "png")
set.seed(123)
```

```{r}
# R version
R.version$version.string
# glmmTMB version
packageVersion("glmmTMB")
```
# Data preparation and exploration

```{r}
# load data
df <- read.csv("data.csv")
# describe data
colnames(df)
dim(df)
df[c("net", "site")]<-lapply(df[c("net", "site")], factor)
summary(df)
```
```{r}
# Calculate mean and standard deviation for numeric columns
mean_values <- sapply(df[, sapply(df, is.numeric)], mean)
std_values <- sapply(df[, sapply(df, is.numeric)], sd)

# Combine results into a data frame
summary_stats <- data.frame(
  Mean = mean_values,
  Standard_Deviation = std_values
)

# Print the summary statistics
print(summary_stats)
```

## Distribution of nflounder
```{r}
num_zeros <- sum(df$nflounder == 0)
total_entries <- length(df$nflounder)
percentage_zeros <- (num_zeros / total_entries) * 100
percentage_zeros
```

```{r}
ggplot(df, aes(nflounder)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  labs(title = "Histogram of nflounder", x = "nflounder")
```
See how many values fall in each category:

```{r}
# Define the bin width
bin_width <- 10

# Define the breaks for the bins
breaks <- seq(min(df$nflounder), max(df$nflounder), by = bin_width)

# Divide the data into bins
bins <- cut(df$nflounder, breaks = breaks, include.lowest = TRUE, right = FALSE)

# Count the number of values in each bin
bin_counts <- table(bins)

# Print the bin counts
print(bin_counts)
```
Lets truncate values above 100 for modelling convenience. 

```{r}
original_nrow <- nrow(df)
df <- subset(df, nflounder <= 100)
removed_nrow <- original_nrow-nrow(df)
conditional_var <- var(df$nflounder, na.rm=TRUE)
conditional_mean <- mean(df$nflounder, na.rm=TRUE)
```

We removed `r removed_nrow` from `r original_nrow`. Let's visualise distribution of nflounder again.

```{r}
# Load necessary libraries
library(ggplot2)

# Create the histogram
ggplot(df, aes(x = nflounder)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black", alpha = 0.7) +
  labs(
    x = "Number of flounder (nflounder)",
    y = "Frequency"
  ) +
  theme_minimal() +  # Use a minimal theme for a clean look
  scale_x_continuous(breaks = seq(0, max(df$nflounder, na.rm = TRUE), by = 10)) +  # Customize x-axis breaks
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))  # Add space at the top of the y-axis
```
As we can see data is still highly overdispersed, the conditional variance (`r conditional_var`) exceeds the conditional mean (`r conditional_mean`). In situations like this negative binomial is an appropriate distribution to use. 

```{r}
ggplot(df, aes(x = factor(year), y = nflounder)) +
  geom_boxplot() +
  # scale_y_log10() +
  labs(x = "year", y = "nflounder", title = "nflounder vs year")
```

## Correlation analysis
```{r}
df_numeric <- df[sapply(df, is.numeric)]
cor_matrix <- cor(df_numeric, use = "complete.obs")
print(cor_matrix)

# Melt the correlation matrix into a long format
cor_matrix_melted <- melt(cor_matrix)

# Plot the heatmap using ggplot2
ggplot(data = cor_matrix_melted, aes(x=Var1, y=Var2, fill=value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Pearson\nCorrelation") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, 
                                   size = 9, hjust = 1), 
        axis.title.x = element_blank(), # Remove x-axis title
        axis.title.y = element_blank(),) +
  
  coord_fixed()
```

## Zero-inflation

```{r}
# Calculate the proportion of values that are 0
zero_proportion <- mean(df$nflounder == 0)
```
The proportion of zeros in nflounder is `r zero_proportion`. 

## Scale variables
We shall scale some of the variables to avoid numerical overflow.

```{r}
var_to_scale <- c("distshore", "area","chlorophyll")
df[, var_to_scale] <- scale(df[, var_to_scale])
```

Consider discarding some of the variables, e.g. `trawl`, in which the proportion of 0's (should be NA's?) `r mean(df$trawl == 0)`.

```{r}
df <- subset(df, select = -trawl)
```
# Data modelling 
## Negative binomial GLM
We fit a negative binomial generalized linear model with log link to the full dataset, excluding site, trawl and net for the moment. This model is useful for count data that exhibit overdispersion (the var exceeds the mean). 
Let \( Y_i \) denote the count response variable for the \( i \)-th observation. The negative binomial distribution for \( Y_i \) is parameterised by the mean \( \lambda_i \) and the dispersion parameter \( \theta \):

\[
Y_i \sim \text{NB}(\lambda_i, \theta)
\]

where the probability mass function is given by:

\[
P(Y_i = k) = \binom{k + \theta - 1}{k} \left( \frac{\theta}{\theta + \lambda_i} \right)^\theta \left( \frac{\lambda_i}{\theta + \lambda_i} \right)^k, \quad k = 0, 1, 2, \ldots
\]

### Log Link Function

The relationship between the mean \( \lambda_i \) and the explanatory variables \( \mathbf{X}_i \) is modeled using a log link function:

\[
\log(\lambda_i) = \mathbf{X}_i \boldsymbol{\beta}
\]

where:
\begin{itemize}
    \item \( \lambda_i \) is the expected count for the \( i \)-th observation.
    \item \( \mathbf{X}_i \) is a vector of explanatory variables for the \( i \)-th observation.
    \item \( \boldsymbol{\beta} \) is a vector of coefficients to be estimated.
\end{itemize}

### Linear Predictor

The linear predictor is given: 
\[
\eta_i = \mathbf{X}_i \boldsymbol{\beta}
\]
where \( \eta_i = \log(\lambda_i) \). Therefore, the model can be rewritten as:

\[
\lambda_i = \exp(\mathbf{X}_i \boldsymbol{\beta})
\]

```{r}
m.glmm.fixed <- glmmTMB(
  nflounder ~ year + lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth, 
  data = df, 
  family = nbinom2()
)
summary(m.glmm.fixed)
```

## Negative Binomial GLMM
Lets add random effect of site and net to the best model. This is an additional term that accounts for unobserved heterogeneity in the data. Random effects are used when data are collected in clusters or groups, and there may be variability between these groups that is not captured by the observed explanatory variables.

### Random Effects

The random effects \( u_j \) are assumed to follow a normal distribution with mean zero and variance \( \sigma^2_u \):

\[
u_j \sim \mathcal{N}(0, \sigma^2_u)
\]

### Linear Predictor

The linear predictor is given by:

\[
\eta_{ij} = \mathbf{X}_{ij} \boldsymbol{\beta} + u_j
\]

where \( \eta_{ij} = \log(\lambda_{ij}) \). Therefore, the model can be rewritten as:

\[
\lambda_{ij} = \exp(\mathbf{X}_{ij} \boldsymbol{\beta} + u_j)
\]

```{r}
# a model of crossed random effects (site and net)
m.glmm.random1 <- glmmTMB(
  nflounder ~ year + lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + (1 | site) + (1 | net), 
  data = df, 
  family = nbinom2()
)
summary(m.glmm.random1)

# alternatively we consider a model of nested random effects (net within a site since nets were reused across sites)

m.glmm.random2 <- glmmTMB(
    nflounder ~ year + lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + (1 | site/net), 
    data = df, 
    family = nbinom2()
)
summary(m.glmm.random2)

# year|site for a model with random variation in slopes through years across sites
m.glmm.random3 <- glmmTMB(
    nflounder ~ lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + (year | site) + (1|net), 
    data = df, 
    family = nbinom2()
)
summary(m.glmm.random3)
```

Lets compare fixed and random effect models.

```{r}
anova(m.glmm.fixed, m.glmm.random1, m.glmm.random2, m.glmm.random3)
```

We choose the model with lowest AIC which is `m.glmm.random2`, a model of nested random effects (net within a site).

## Zero-inflated poisson GLMM

Now lets try a zero-inflated Poisson model with a single zero inflation parameter applying to all observations using ziformula~1.

```{r}
m.glmm.random.zero <- glmmTMB(
    nflounder ~ year + lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + (1 | site/net), 
    data = df, 
    ziformula=~1,
    family = poisson()
)
summary(m.glmm.random.zero)
```
This model has higher AIC, hence we prefer binomial models.

## Truncated Negative Binomial GLMM (Hurdle Model).

In contrast to zero-inflated models, hurdle models treat zero-count and nonzero outcomes as two completely separate categories, rather than treating the zero-count outcomes as a mixture of structural and sampling zeros.

```{r}
m.glmm.random2.hurdle <- update (m.glmm.random2,
                                 data=df,
                                 ziformula=~.,
                                 family=truncated_nbinom2())
summary(m.glmm.random2.hurdle)
```
Lets compare the models.
```{r}
anova(m.glmm.random2, m.glmm.random2.hurdle)
```

Model `m.glmm.random2.hurdle` is significantly better. From now on we continue with the negative binomial GLMM Hurdle model.
This model assumes the following:

\begin{itemize}
    \item Count data: The response variable \(Y\) represents counts of events or occurrences.
    \item Conditional distribution: The counts are assumed to follow a Negative Binomial distribution.
    \item Random effects: The model incorporates random effects to account for unobserved heterogeneity among groups or individuals.
    \item Excess zeros: The excess zeros in the data are modeled separately from the counts using a hurdle model approach.
\end{itemize}

### Parameterization

Let \(Y_{ij}\) denote the count response variable for the \(i\)-th observation in the \(j\)-th group or individual. The model is parameterised as follows:

\begin{enumerate}
    \item \textbf{Count Model}:
    \[
    Y_{ij} \sim \text{NegBin}(\lambda_{ij}, \theta)
    \]
    where \(\lambda_{ij}\) is the mean of the Negative Binomial distribution and \(\theta\) is the dispersion parameter.
    
    \item \textbf{Zero-Inflation Model}:
    \[
    Z_{ij} \sim \text{Bernoulli}(\pi_{ij})
    \]
    where \(Z_{ij}\) represents a binary indicator for excess zeros and \(\pi_{ij}\) is the probability of observing excess zeros.
\end{enumerate}

### Link Function

The relationship between the mean count \(\lambda_{ij}\) and the predictors \(X_{ij}\) is modeled using a log link function:

\[
\log(\lambda_{ij}) = X_{ij}\beta + u_j
\]

where \(X_{ij}\) is a vector of fixed effect predictors, \(\beta\) is a vector of fixed effect coefficients, and \(u_j\) represents the random effect for the \(j\)-th group or individual.

### Interpretation

The Negative Binomial GLMM estimates the effects of the predictors on both the count of events and the presence of excess zeros. The fixed effect coefficients (\(\beta\)) quantify the impact of the predictors on the mean count (\(\lambda_{ij}\)), while the random effects (\(u_j\)) capture the variability among groups or individuals.

## Post-model-fitting procedure. 
### Residuals
```{r}
plot(simulateResiduals(m.glmm.random2.hurdle))
```

### Estimated marginal means

Lets consider marginal means between net types.

```{r}
m.glmm.random2.hurdle.net <- glmmTMB(
    nflounder ~ year + lat + long + distshore + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + net + (1 | site/net), 
    data = df, 
    ziformula=~.,
    family=truncated_nbinom2()
)
summary(m.glmm.random2.hurdle.net)

emmeans(m.glmm.random2.hurdle.net,"net")
```

### Variable selection with drop1

Refit the model with various terms dropped.

```{r}
system.time(m.glmm.random2.hurdle.d1 <- drop1(m.glmm.random2.hurdle,test="Chisq"))
print(m.glmm.random2.hurdle.d1)
```
Lets remove some of the variables (bod, nh3, po4).

```{r}
final <- glmmTMB(nflounder ~ year + lat + long + distshore + chlorophyll + tempavg + tempstdev + sal + depth + (1 | site/net), 
    data = df, 
    ziformula=~.,
    family=truncated_nbinom2())
summary(final)
```

## Conclusion

The response variable nflounder is modeled using a truncated negative binomial distribution with a log link function (Hurdle). 

Truncated Negative Binomial model is a two part model, where the zero-inflation model estimates the presence/absence; the conditional model estimates the abundance. 

The predictors in the both parts of the model include year, lat, long, distshore, chlorophyll, tempavg, tempstdev, sal, and depth, along with a random intercept for site/net. 

### Random effect

#### Conditional Model:
- net:site Group (Intercept): The variability in the response variable nflounder due to the grouping of nets within sites is relatively low, with a variance of 0.08427 and a standard deviation of 0.2903.
- site Group (Intercept): The variability in nflounder due to different sites is higher compared to the net:site grouping, with a variance of 0.70174 and a standard deviation of 0.8377.

#### Zero-Inflation Model:
net:site Group (Intercept): The variability in the zero-inflation component attributable to the grouping of nets within sites is moderate, with a variance of 0.1120 and a standard deviation of 0.3347.
site Group (Intercept): The variability in zero-inflation due to different sites is relatively higher, with a variance of 0.4647 and a standard deviation of 0.6817.

### Fixed Effects

#### Conditional Model:
The predictor year is statistically significant (p < 0.001), with a negative coefficient indicating a decrease in nflounder over time.
distshore and salinity also show significance (p < 0.001 and p < 0.05, respectively), with negative coefficients implying a negative association with nflounder.
Other predictors such as latitude, longitude, chlorophyll, average temperature, temperature standard deviation, and depth do not show statistically significant associations with nflounder.

#### Zero-Inflation Model:
Similar to the conditional model, year, latitude, longitude, distshore, and salinity are statistically significant predictors (p < 0.001 or p < 0.05).
temperature standard deviation also shows significance (p < 0.05), but with a negative coefficient, indicating a negative association with the zero-inflation component.

Overal, including random effects for site and net reduces AIC and improves model fit, indicating that accounting for site-specific and net-specific variability is important. Nested random effects (site/net) further improve the model fit slightly compared to separate random effects. The models indicate that both spatial (longitude, distance to shore) and environmental (temperature, salinity, depth) variables significantly influence flounder counts. Random effects for site and net are crucial for capturing variability in the data, suggesting that there are site-specific and gear-specific influences on flounder counts. The year variable shows a strong temporal trend, which could indicate changes in flounder population over time.

## References

Brooks ME, Kristensen K, van Benthem KJ, Magnusson A, Berg CW, Nielsen A, Skaug HJ, Maechler M, Bolker BM (2017). “glmmTMB Balances Speed and Flexibility Among Packages for Zero-inflated Generalized Linear Mixed Modeling.” The R Journal, 9(2), 378–400. doi:10.32614/RJ-2017-066.
 