---
title: "Statistical analysis of the effect of environmental variables on abundance of flounder"
author: "Olga Lyashevska"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup}
# load packages 
packages <- c("ggplot2", "MASS", "mgcv", "rmarkdown", "tinytex", "lme4")
lapply(packages, library, character.only = TRUE)
knitr::opts_chunk$set(fig.path = "figure/", dev = "png")
```

# Data preparation and exploration

```{r}
# load data
df <- read.csv("data.csv")
# describe data
colnames(df)
dim(df)
df[c("net", "site")]<-lapply(df[c("net", "site")], factor)
summary(df)
```

## Distribution of nflounder

```{r}
ggplot(df, aes(nflounder)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  labs(title = "Histogram of nflounder", x = "nflounder")
```
See how many values fall in each category:

```{r}
# Define the bin width
bin_width <- 10

# Define the breaks for the bins
breaks <- seq(min(df$nflounder), max(df$nflounder), by = bin_width)

# Divide the data into bins
bins <- cut(df$nflounder, breaks = breaks, include.lowest = TRUE, right = FALSE)

# Count the number of values in each bin
bin_counts <- table(bins)

# Print the bin counts
print(bin_counts)
```
Lets truncate values above 100 for modelling convenience. 

```{r}
original_nrow <- nrow(df)
df <- subset(df, nflounder <= 100)
removed_nrow <- original_nrow-nrow(df)
conditional_var <- var(df$nflounder, na.rm=TRUE)
conditional_mean <- mean(df$nflounder, na.rm=TRUE)
```

We removed `r removed_nrow` from `r original_nrow`. Let's visualise distribution of nflounder again.
```{r}
ggplot(df, aes(nflounder)) + 
  geom_histogram(binwidth = 10, fill = "blue", color = "black") +
  labs(title = "Histogram of nflounder", x = "nflounder")
```
As we can see data is still highly overdispersed, the conditional variance (`r conditional_var`) exceeds the conditional mean (`r conditional_mean`). In situations like this negative binomial is an appropriate distribution to use. 

```{r}
ggplot(df, aes(x = factor(year), y = nflounder)) +
  geom_boxplot() +
  # scale_y_log10() +
  labs(x = "year", y = "nflounder", title = "nflounder vs year")
```

## Correlation analysis
```{r}
df_numeric <- df[sapply(df, is.numeric)]
cor_matrix <- cor(df_numeric, use = "complete.obs")
print(cor_matrix)
```

## Zero-inflation

```{r}
# Calculate the proportion of values that are 0
zero_proportion <- mean(df$nflounder == 0)
```
The proportion of zeros in nflounder is `r zero_proportion`. 

## Scale variables
We shall scale some of the variables to avoid numerical overflow.

```{r}
var_to_scale <- c("distshore", "trawl", "area","chlorophyll")
df[, var_to_scale] <- scale(df[, var_to_scale])
```

# Data modelling 
## Negative binomial GLM
Lets fit a negative binomial generalized linear model for full dataset, excluding site and net for the moment.

```{r}
m.glm <- glm.nb(nflounder ~. - site - net  , data = df)
summary(m.glm)
```
Choose a model by AIC in a Stepwise Algorithm

```{r results='hide'}
m.glm.stp<-stepAIC(m.glm, direction="both")
```

```{r}
summary(m.glm.stp)
```

Likelihood ratio test, p-value tells if the likelihoods can be considered different. Non-significant p-value, use model with smaller df.

```{r}
anova(m.glm, m.glm.stp)
```
## Negative Binomial GLMM
Add random effect of site and net to the best model.

<!-- ```{r} -->
<!-- # library(lme4) -->
<!-- # m.glmm <- glmer.nb(nflounder ~ year + lat + long + distshore + trawl + area + chlorophyll + tempavg + tempstdev + sal + bod + nh3 + po4 + depth + (1|site)+(1|net), data = df) -->
<!-- # summary() -->
<!-- ``` -->

## Zero-inflated negative binomial GLM
<!-- https://fukamilab.github.io/BIO202/04-C-zero-data.html#zero-inflated_negative_binomial_glm -->
## Zero-inflated negative binomial GLMM

## Negative binomial GAM
<!-- m.gam <- gam(nflounder ~ net + year + long + trawl + area + s(chlorophyll) +  -->
<!--               tempavg + tempstdev + s(sal) + nh3 + depth +  -->
<!--               distshore_cube + area_cube + s(bod),  -->
<!--               data = df, family = "nb", method = "REML", select=TRUE) -->

<!-- summary(m.gam) -->

<!-- png("m.gam.png", width = 900, height = 900, res = 120) -->
<!-- par(mfrow = c(2,2)) -->
<!-- gam.check(m.gam) -->
<!-- dev.off() -->

<!-- png("m.gam.smooth.png", width = 900, height = 900, res = 120) -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(m.gam) -->
<!-- dev.off() -->

<!-- # drop non-significant variables -->
<!-- m.gam.sig <- gam(nflounder ~ net + year + long + area + s(chlorophyll) +  -->
<!--                tempstdev + s(sal) + depth +  -->
<!--                distshore_cube + area_cube + s(bod),  -->
<!--              data = df, family = "nb", method = "REML", select=TRUE) -->

<!-- summary(m.gam.sig) -->

<!-- png("m.gam.sig.png", width = 900, height = 900, res = 120) -->
<!-- par(mfrow = c(2,2)) -->
<!-- gam.check(m.gam.sig) -->
<!-- dev.off() -->

<!-- png("m.gam.sig.smooth.png", width = 900, height = 900, res = 120) -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(m.gam.sig) -->
<!-- dev.off() -->

## Negative binomial GAMM
Consider adding random effect of site  and net
<!-- # Generalised Additive Mixed Modelling -->
<!-- m.gamm <- gamm(nflounder ∼ nflounder ~ net + year + long + area + s(chlorophyll) +  -->
<!--                  tempstdev + s(sal) + depth + distshore_cube + area_cube + s(bod),  -->
<!--                random = list(site = ∼ 1), data = df) -->

<!-- https://r.qcbs.ca/workshop08/book-en/introduction-to-generalized-additive-mixed-models-gamms.html -->
<!-- m.gamm <- gamm(nflounder ~ net + year + long + area + s(chlorophyll) + -->
<!--                  tempstdev + s(sal) + depth + distshore_cube + area_cube + s(bod), -->
<!--                random = list(site = ~ 1), data = df, family = "nb", method = "REML", select = TRUE) -->




<!-- # model diagnostics -->
<!-- # there are a few outliers, check -->
<!-- # residuals vs fitted --> linearity assumptions -->
<!-- # qqplot -> normality -->
<!-- # scale-location --> homogeneity variance -->
<!-- # residuals vs leverage --> influential points  -->
<!-- png("m.glm.stp.33.png", width = 900, height = 900, res = 120) -->
<!-- par(mfrow=c(2,2)) -->
<!-- plot(m.glm.stp) -->
<!-- dev.off() -->

<!-- m.glm.res = residuals(m.glm) -->
<!-- We now plot the residual against the observed values of the variables -->
<!-- plot(m.glm, m.glm.res) -->
<!-- plot(m.glm, m.glm.res)+ -->
<!--   ylab("residuals") + xlab("variable") -->
<!-- <return> -->
